---
title: "Early life effects on health in OA"
output: html_document
date: "2025-05-21"
---

#load packages needed for analyses
```{r}
library(tidyverse)
library(mice)
library(ggpubr)
library(ggmap)
library(nlWaldTest)
library(nationalparkcolors)
pal <- park_palette("Badlands")
```

#or load exact package versions using:
```{r}
install.packages("renv") #if not installed
renv::restore()
```

#load data
```{r}
OA_data<-read.delim("27Mar25_OA_merged.txt")
```


## Generate early life PC scores
```{r}
# Variable Groups
parent_occupation <-   c("early_father_occupation___crop",
                         "early_father_occupation___subsist_crop",
                         "early_father_occupation___forage",
                         "early_father_occupation___wage",
                         "early_mother_occupation___crop",
                         "early_mother_occupation___subsist_crop",
                         "early_mother_occupation___forage",
                         "early_mother_occupation___wage")

hh_item <- c( "early_hh_item___moto",
              "early_hh_item___bicycle",
              "early_hh_item___generator",
              "early_hh_item___blowpipe",
              "early_hh_item___machete",
              "early_hh_item___flip_phone",
              "early_hh_item___smart_phone",
              "early_hh_item___chainsaw",
              "early_hh_item___weedwacker",
              "early_hh_item___car",
              "early_hh_item___tv",
              "early_hh_item___gas_stove")

diet <- c("early_store",
          "early_oil_sugar",
          "early_wild_meat")

early_misc <- c("early_school",
                "birth_place_hosp",
                "early_vill_access_road",
                "early_clean_water"  )

housing <- c( "early_house_type___traditional",
               "early_house_type___stone",
               "early_house_type___wood")


# List of variables to check for completeness
variables_to_check <- c(parent_occupation, diet, housing, early_misc, hh_item)

# how many variables are missing for a given person
x<-(apply( OA_data[,variables_to_check] ,1,function(x) length(which(is.na(x))) ))
table(x)

# Remove observations with missing data for variables_to_check (allow 1 missing data point)
OA_data_complete <- OA_data[which(x<2), variables_to_check]

#Impute missing data
OA_data_complete_imputed <- complete(mice(OA_data_complete, method ="lasso.norm"))

# Perform PCA
pca_result <- prcomp(OA_data_complete_imputed, center = TRUE, scale. = TRUE)

#Extract skree plot 
s <- summary(pca_result)
dat <- data.frame(
  component = factor(1:length(s$sdev), labels=paste0("PC", 1:length(s$sdev))),
  var_explained = s$sdev^2/sum(s$sdev^2)
)
skree_plot<-ggplot(dat[1:10,], aes(y=var_explained)) + 
  geom_line(aes(x=component, group=1)) + 
  geom_point(aes(x=component)) + 
  theme_bw(13) + 
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1))+
  ylab('Percent variance explained');skree_plot

# Extract the PC scores
OA_data$Early_urban_PC1<-NA
OA_data$Early_urban_PC1[which(x<2)]<-pca_result$x[,1]
OA_data$Early_urban_PC2<-NA
OA_data$Early_urban_PC2[which(x<2)]<-pca_result$x[,2]
```

### Make the barplot of loadings for PC1
```{r} 
# Extract PC1 loadings
loadings_df <- as.data.frame(pca_result$rotation[,1])

names(loadings_df)<-'loadings'
loadings_df$variable<-rownames(loadings_df)
loadings_df$group<-NA
loadings_df$group[which(loadings_df$variable %in% diet)]<-'diet'
loadings_df$group[which(loadings_df$variable %in% parent_occupation)]<-'parent occupation'
loadings_df$group[which(loadings_df$variable %in% early_misc)]<-'misc'
loadings_df$group[which(loadings_df$variable %in% hh_item)]<-'household items'
loadings_df$group[which(loadings_df$variable %in% housing)]<-'housing'

# Mapping for cleaner labels
clean_labels <- c(
  "early_father_occupation___crop" = "Father Crop",
  "early_father_occupation___subsist_crop" = "Father Subsistence Crop",
  "early_father_occupation___forage" = "Father Forage",
  "early_father_occupation___wage" = "Father Wage",
  "early_mother_occupation___crop" = "Mother Crop",
  "early_mother_occupation___subsist_crop" = "Mother Subsistence Crop",
  "early_mother_occupation___forage" = "Mother Forage",
  "early_mother_occupation___wage" = "Mother Wage",
  "early_hh_item___moto" = "Motorcycle",
  "early_hh_item___bicycle" = "Bicycle",
  "early_hh_item___generator" = "Generator",
  "early_hh_item___blowpipe" = "Blowpipe",
  "early_hh_item___machete" = "Machete",
  "early_hh_item___flip_phone" = "Flip Phone",
  "early_hh_item___smart_phone" = "Smart Phone",
  "early_hh_item___chainsaw" = "Chainsaw",
  "early_hh_item___weedwacker" = "Weedwacker",
  "early_hh_item___car" = "Car",
  "early_hh_item___tv" = "TV",
  "early_hh_item___gas_stove" = "Gas Stove",
  "early_house_type___traditional" = "Traditional House",
  "early_house_type___stone" = "Stone House",
  "early_house_type___wood" = "Wood House",
  "early_store" = "Store Bought Food",
  "early_oil_sugar" = "Oil & Sugar Use",
  "early_wild_meat" = "Wild Meat Consumption",
  "early_school" = "Nearby School",
  "highest_education_stage" = "Highest Education",
  "birth_place_hosp" = "Born in Hospital",
  "early_vill_access_road"= "Village accessible by road",
  "early_clean_water" = "Natural water source"
)

# Convert to dataframe for plotting
loadings_df <- loadings_df %>%
  mutate(variable = recode(variable, !!!clean_labels))

traditional_variables<-c("Traditional House", "Machete", "Blowpipe", "Wild Meat Consumption", "Mother Forage", "Father Forage", "Mother Crop", "Father Crop", "Mother Subsistence Crop", "Father Subsistence Crop", "Natural water source")
loadings_df$traditional_group<- ifelse(loadings_df$variable %in% traditional_variables, "Traditional", "Urban")

# Create a grouped variable
loadings_df <- loadings_df %>%
  mutate(grouped_x = interaction(variable, group))

# Order by value within each category
loadings_df <- loadings_df %>%
  arrange(group, desc(loadings)) %>%  # Sort within each category
  mutate(grouped_x = factor(grouped_x, levels = unique(grouped_x)))  

#Rename the variable groups
loadings_df$group<-factor(loadings_df$group, levels=c("diet", "household items", "housing", "misc", "parent occupation"), labels=c("diet", "household items", "housing", "community attributes", "parent livelihood"))

PC1_bar<-ggplot(loadings_df, aes(x = grouped_x, y = loadings, fill = group, alpha=traditional_group)) +
  geom_bar(stat = "identity", color="black") +
  coord_flip() +
  labs(title = "PC1 Loadings", x = "Early life variables", y = "Loading", fill="Variable Groups") +
  theme_minimal() +
     guides(alpha = "none") + 
  scale_alpha_manual(values= c(0.5, 1)) +
  scale_x_discrete(labels = function(x) sub("\\..*", "", x)) +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text( hjust=0.5), legend.position="right") +
  scale_fill_manual(values=c(pal[1], "grey", pal[2], pal[3], pal[4]))+
  guides(fill = guide_legend(reverse = TRUE)); PC1_bar
```

#boxplot of PC1 by village
```{r}
OA_data$village_id<-as.factor(OA_data$village_id)
OA_data_plot<-OA_data %>%
  filter(!is.na(village_id)) %>%
  filter(!is.na(Early_urban_PC1)) %>%
  group_by(village_id) %>%
  mutate(mean_PC1= mean(Early_urban_PC1, na.rm=T)) %>%
  ungroup()

PC1_by_village<-ggplot(OA_data_plot, aes(x=reorder(village_id, -mean_PC1), y=Early_urban_PC1)) +
  theme_bw() +
  geom_hline(yintercept=0, linetype="dashed")+
  geom_boxplot(outlier.size=1) +
  ylab("Early life urbanicity: PC1") +
  xlab("Village ID"); PC1_by_village
```

### standardize variables
```{r} 
# recode sex
OA_data_std <- OA_data %>%
  mutate(sex = as.character(sex)) %>%
  mutate(sex = case_when(
    sex == "1" ~ "Male",
    sex == "0" ~ "Female",
    TRUE ~ sex  # This keeps any existing character values
  ))

Body_Height <- c("height_standing", "height_hip", "height_knee")

Body_Press <- c("waist_circum", "weight", "BMI", "BRI", "body_fat", 
                "waist_hip","sys_bp2", "dias_bp2")

Blood_Bio <- c("total_chol", "ldl", "hdl",  "triglyc")
Binary<-c('obese','hypertension')

standardize <- function(x) {
  return((x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))
}
OA_merged_std <- OA_data_std %>%
  mutate(across(all_of(c("age","urb_score",Body_Height, Body_Press, Blood_Bio)), 
                .fns = ~ standardize(.x), 
                .names = "{.col}_std"))
OA_merged<-subset(OA_merged_std,age>17)
OA_merged$Early_PC1_std<- scale(OA_merged$Early_urban_PC1)

# sanity check the cleaned variables (before standardization)
tmp<-OA_merged[,c(Body_Height,Body_Press,Blood_Bio,'rid','sex')] %>%
  pivot_longer(!rid & !sex, names_to = "outcome", values_to = "value")

tmp$outcome <- recode(tmp$outcome,
  height_standing = "Standing height",
  height_hip = "Hip height",
  height_knee = "Knee height",
  waist_circum = "Waist circumference",
  weight = "Body weight",
  BMI = "BMI",
  BRI = "BRI",
  body_fat = "Body fat percentage",
  waist_hip = "Waist to hip ratio",
  sys_bp2 = "Systolic BP",
  dias_bp2 = "Diastolic BP",
  total_chol = "Total cholesterol",
  ldl = "LDL",
  hdl = "HDL",
  triglyc = "Triglycerides"
)

tmp$outcome <- factor(tmp$outcome, levels = c("Standing height","Hip height","Knee height","Waist circumference","Body weight","BMI","BRI","Body fat percentage","Waist to hip ratio","Systolic BP","Diastolic BP","Total cholesterol","LDL","HDL","Triglycerides"))

MvF_density<-ggplot(tmp, aes(x=value,fill=sex)) + geom_density(alpha=0.5)+ facet_wrap(~outcome,scales='free')+ theme(legend.position="bottom") +
  theme_bw() +
  xlab("Biomarker value");MvF_density
```


### Make some maps - current sampling locations and current urbanicity
```{r} 
# Prep the map
medianlat <- median(OA_merged$lat, na.rm = TRUE)
medianlong <- median(OA_merged$long, na.rm = TRUE)

register_google(Sys.getenv("GOOGLE_API_KEY"))

s <- "element:geometry%7Ccolor:0xf5f5f5&style=element:labels%7Cvisibility:off&style=element:labels.icon%7Cvisibility:off&style=element:labels.text.fill%7Ccolor:0x616161&style=element:labels.text.stroke%7Ccolor:0xf5f5f5&style=feature:administrative%7Celement:geometry%7Cvisibility:off&style=feature:administrative.country%7Celement:geometry.stroke%7Ccolor:0x000000%7Cvisibility:on&style=feature:administrative.land_parcel%7Cvisibility:off&style=feature:administrative.land_parcel%7Celement:labels.text.fill%7Ccolor:0xbdbdbd&style=feature:administrative.neighborhood%7Cvisibility:off&style=feature:poi%7Cvisibility:off&style=feature:poi%7Celement:geometry%7Ccolor:0xeeeeee&style=feature:poi%7Celement:labels.text.fill%7Ccolor:0x757575&style=feature:poi.park%7Celement:geometry%7Ccolor:0xe5e5e5&style=feature:poi.park%7Celement:labels.text.fill%7Ccolor:0x9e9e9e&style=feature:road%7Cvisibility:off&style=feature:road%7Celement:geometry%7Ccolor:0xffffff&style=feature:road%7Celement:labels.icon%7Cvisibility:off&style=feature:road.arterial%7Celement:labels.text.fill%7Ccolor:0x757575&style=feature:road.highway%7Celement:geometry%7Ccolor:0xdadada&style=feature:road.highway%7Celement:labels.text.fill%7Ccolor:0x616161&style=feature:road.local%7Celement:labels.text.fill%7Ccolor:0x9e9e9e&style=feature:transit%7Cvisibility:off&style=feature:transit.line%7Celement:geometry%7Ccolor:0xe5e5e5&style=feature:transit.station%7Celement:geometry%7Ccolor:0xeeeeee&style=feature:water%7Celement:geometry%7Ccolor:0xc9c9c9&style=feature:water%7Celement:labels.text.fill%7Ccolor:0x9e9e9e&size=480x360"
c(medianlong, medianlat)

map <- get_googlemap(center=c(102,4), zoom = 7, scale = 1, style = s)
m <- ggmap(map)
```

#map of sampling locations
```{r}
sample_local_plot<- m +
 geom_point(data = OA_merged, aes(x = long, y = lat), fill=pal[2], color="black", alpha = 0.2, size=3, shape=21) +
  xlab("Longitude") +
  ylab("Latitude") +
  xlim(100.5, 103.5) +
  ylim(2, 6.5) +
  theme_bw(13); sample_local_plot
```

#map of current urbanicity
```{r} 
Current_urb_by_village<- OA_merged %>%
  group_by(village_id) %>%
  summarize(
    mean_lat= mean(lat),
    mean_long=mean(long),
    mean_urb = mean(urb_score, na.rm=T), 
    count = n() )

Current_urb_by_village<-subset(Current_urb_by_village,count>9)

current_urb_plot<- m +
 geom_point(data = Current_urb_by_village, aes(x = mean_long, y = mean_lat, fill=mean_urb, size=count), color="black", alpha = 0.8, shape=21) +
  labs(x="Longitude", y="Latitude", size="Sample \n  size", fill="    Adult \n urbanicity \n    score") +
   scale_fill_gradient(low = "#5CB9F4", high = pal[3]) +  
  xlim(100.5, 103.5) +
  ylim(2, 6.5) +
  theme_bw(13); current_urb_plot
```


#Early life urbanicity by birth year
```{r}
#Make color scale based on entire sample
OA_village_level <- OA_merged %>%
  group_by(village_id) %>%
  summarise(
    lat = mean(lat, na.rm = TRUE),
    long = mean(long, na.rm = TRUE),
    avg_Early_urban_PC1 = mean(Early_urban_PC1, na.rm = TRUE),
    avg_Early_urban_PC2 = mean(Early_urban_PC2, na.rm = TRUE),
    avg_urb_score = mean(urb_score, na.rm = TRUE),
    count = n()  # Count the number of rows in each group
  )

#find min and max for the whole dataset:
min(OA_village_level$avg_Early_urban_PC1, na.rm=T) #-2.4
max(OA_village_level$avg_Early_urban_PC1, na.rm=T) #3.96

OA_merged$birth_year <- year(OA_merged$date_of_birth)
OA_village_level_old <- OA_merged %>%
  filter(birth_year <= 1975) %>%
  group_by(village_id) %>%
  summarise(
    lat = mean(lat, na.rm = TRUE),
    long = mean(long, na.rm = TRUE),
    avg_Early_urban_PC1 = mean(Early_urban_PC1, na.rm = TRUE),
    avg_Early_urban_PC2 = mean(Early_urban_PC2, na.rm = TRUE),
    avg_urb_score = mean(urb_score, na.rm = TRUE),
    count = n()  # Count the number of rows in each group
  )
OA_village_level_old<-subset(OA_village_level_old,!is.na(avg_Early_urban_PC1))


PC1_plot_1975 <- m +
  geom_point(data = OA_village_level_old, aes(x = long, y = lat, fill = avg_Early_urban_PC1, size = count), alpha = 0.8, shape=21, color="black") +
  scale_fill_gradient(low = "#5CB9F4", high = pal[3], 
                       name = "Early life PC1", 
                       limits= c(-2.5, 6.8),
                       breaks = c(-1.8, 5.8),
                       labels = c("More traditional", "More urban")) +  # Match labels to breaks
  scale_size_continuous(range = c(2,5), name = "Count") +  # Larger circles
  xlab("Longitude") +
  ylab("Latitude") +
  xlim(100.5, 104) +
  ylim(2, 6.5) +
  theme_bw(13) +
  ggtitle("Born before 1975") +
  theme(legend.position = "right", plot.title=element_text(hjust=0.5)); PC1_plot_1975

ggsave("EarlyLifePC1_map_1975.pdf", PC1_plot_1975)


#after 2000
OA_village_level_young <- OA_merged %>%
  filter(birth_year >= 2000) %>%
  group_by(village_id) %>%
  summarise(
    lat = mean(lat, na.rm = TRUE),
    long = mean(long, na.rm = TRUE),
    avg_Early_urban_PC1 = mean(Early_urban_PC1, na.rm = TRUE),
    avg_Early_urban_PC2 = mean(Early_urban_PC2, na.rm = TRUE),
    avg_urb_score = mean(urb_score, na.rm = TRUE),
    count = n()  # Count the number of rows in each group
  )
OA_village_level_young<-subset(OA_village_level_young,!is.na(avg_Early_urban_PC1))

PC1_plot_2000 <- m +
  geom_point(data = OA_village_level_young, aes(x = long, y = lat, fill = avg_Early_urban_PC1, size = count), alpha = 0.8, shape=21, color="black") +
  scale_fill_gradient(low = "#5CB9F4", high = pal[3], 
                       name = "Early life PC1", 
                       limits= c(-2.5, 6.8),
                       breaks = c(-1.8, 5.8),    # Define breaks at the extremes of the scale
                       labels = c("More traditional", "More urban")) +  # Match labels to breaks
  scale_size_continuous(range = c(2,5), name = "Count") +  # Larger circles
  xlab("Longitude") +
  ylab("Latitude") +
  xlim(100.5, 104) +
  ylim(2, 6.5) +
  theme_bw(13) +
  ggtitle("Born after 2000") +
  theme(legend.position = "right", plot.title=element_text(hjust=0.5)); PC1_plot_2000


EarlyLifePC1_multimap2<-ggarrange(PC1_plot_1975, PC1_plot_2000, common.legend=T, legend="right"); EarlyLifePC1_multimap2
```

             #####################################################################################
             ### Run the main models for early life PC1, current urbanicity & the interaction ###
             #####################################################################################

```{r} 
# List of outcome variables - std
Body_Height <- c("height_standing_std", "height_hip_std", "height_knee_std")

Body_Press <- c("waist_circum_std", "weight_std", "BMI_std", "BRI_std", "body_fat_std", 
                "waist_hip_std","sys_bp2_std", "dias_bp2_std")

Blood_Bio <- c("total_chol_std", "ldl_std", "hdl_std", "triglyc_std")
Binary<-c('obese','hypertension')

all_outcomes<-c(Body_Height,Body_Press,Blood_Bio,Binary)
results_pval<-matrix(nrow=length(all_outcomes),ncol=5)
results_est<-matrix(nrow=length(all_outcomes),ncol=5)
results_CI1<-matrix(nrow=length(all_outcomes),ncol=5)
results_CI2<-matrix(nrow=length(all_outcomes),ncol=5)
results_N<-matrix(nrow=length(all_outcomes),ncol=1)

for (i in 1:length(c(Body_Height,Body_Press,Blood_Bio))){
OA_merged$tmp_outcome<-OA_merged[,which(names(OA_merged) %in% all_outcomes[i])]
x<-(lm( tmp_outcome ~ sex + age_std + urb_score_std * Early_PC1_std ,data=OA_merged))
results_pval[i,1:5]<-summary(x)$coefficients[c(2:6),4]
results_est[i,1:5]<-summary(x)$coefficients[c(2:6),1]
results_CI1[i,1:5]<-confint(x)[c(2:6),1]
results_CI2[i,1:5]<-confint(x)[c(2:6),2]
results_N[i,1]<-summary(x)$df[2] + summary(x)$df[1]
}

for (i in 1:length(Binary)){
OA_merged$tmp_outcome<-OA_merged[,which(names(OA_merged) %in% Binary[i])]
x<-(glm( tmp_outcome ~ sex + age_std + urb_score_std * Early_PC1_std ,data=OA_merged,family='binomial'))
results_pval[i+15,1:5]<-summary(x)$coefficients[c(2:6),4]
results_est[i+15,1:5]<-summary(x)$coefficients[c(2:6),1]
results_CI1[i+15,1:5]<-confint(x)[c(2:6),1]
results_CI2[i+15,1:5]<-confint(x)[c(2:6),2]
results_N[i+15,1]<-summary(x)$df[2] + summary(x)$df[1]
}

results_pval<-as.data.frame(results_pval)
names(results_pval)<-c('sex','age','urb_score','early_PC1','int')
results_est<-as.data.frame(results_est)
names(results_est)<-c('sex','age','urb_score','early_PC1','int')
results_N<-as.data.frame(results_N)
results_est$outcome<-all_outcomes
results_pval$outcome<-all_outcomes
results_N$outcome<-all_outcomes
```


#export model results
```{r}
estimates<-pivot_longer(data=results_est, cols=1:5, names_to= "test", values_to="estimate")
pvalues<- pivot_longer(data=results_pval, cols=1:5, names_to= "test", values_to="pvalue")

merged<- merge(estimates, pvalues)
merged$FDR<- p.adjust(merged$pvalue,method='BH')
merged2<-merge(merged, results_N)
colnames(merged2)<- c("outcome", "predictor", "estimate", "pvalue", "FDR", "nobs")

merged2$outcome <- recode(merged2$outcome,
  height_standing_std = "Standing height",
  height_hip_std = "Hip height",
  height_knee_std = "Knee height",
  waist_circum_std = "Waist circumference",
  weight_std = "Body weight",
  BMI_std = "BMI",
  BRI_std = "BRI",
  body_fat_std = "Body fat percentage",
  waist_hip_std = "Waist to hip ratio",
  sys_bp2_std = "Systolic BP",
  dias_bp2_std = "Diastolic BP",
  total_chol_std = "Total cholesterol",
  ldl_std = "LDL",
  hdl_std = "HDL",
  triglyc_std = "Triglycerides"
)

merged2$predictor <- recode(merged2$predictor,
  early_PC1 = "early life PC1",
  int = "interaction",
  urb_score = "current urbanicity")

merged2$significant<-'ns'
merged2$significant[which(merged2$pvalue<0.05)]<-'p<0.05'
merged2$significant[which(merged2$FDR<0.1)]<-'FDR<0.1'
```


```{r} 
# run some additional models for BP
OA_merged$ethnicity_negrito<-ifelse(OA_merged$ethnicity2=='Semang',1,0)
OA_merged$tobacco2<-ifelse(OA_merged$tobacco=='never',0,1)
OA_merged$tobacco_leaf2<-ifelse(OA_merged$tobacco_leaf=='never',0,1)
OA_merged$betel2<-ifelse(OA_merged$betel=='never',0,1)

summary(lm( sys_bp2_std ~ sex + age_std + urb_score_std * Early_PC1_std +tobacco2 +tobacco_leaf2 ,data=OA_merged))
summary(lm( dias_bp2_std ~ sex + age_std + urb_score_std * Early_PC1_std +tobacco2 +tobacco_leaf2 ,data=OA_merged))

```

### plot an example: Body fat
```{r} 
OA_merged$early_PC_split<-ifelse(OA_merged$Early_PC1_std<0,'rural','urban')
tmp<-subset(OA_merged,!is.na(early_PC_split) )

data_summary <- function(x) {
   m <- mean(x)
   ymin <- m-sd(x)
   ymax <- m+sd(x)
   return(c(y=m,ymin=ymin,ymax=ymax))
}

Body_fat_EL_sexstrat<-ggplot(tmp, aes(x=sex,y=body_fat,fill=early_PC_split)) + 
  geom_violin()+ stat_summary(fun.data=data_summary,position=position_dodge(0.9))+
  theme_bw(12) + 
  theme(legend.position="bottom", legend.title = element_text(size = 12),
    legend.text = element_text(size = 8), axis.text=element_text(color="black")) + 
  scale_color_manual(values=pal[1:2], labels=c("rural" = "rural \n(PC1 < 0)", "urban"= "urban \n(PC1 > 0)")) +
  guides(fill = guide_legend(title = "Early life urbanicity")) +
  scale_fill_manual(values=c("gray70", "white")) +
  xlab("") +
  ylab("Body fat percentage");Body_fat_EL_sexstrat
```


### plot an example: Knee height
```{r} 
OA_merged$early_PC_split<-ifelse(OA_merged$Early_PC1_std<0,'rural','urban')
tmp<-subset(OA_merged,!is.na(early_PC_split) )

data_summary <- function(x) {
   m <- mean(x)
   ymin <- m-sd(x)
   ymax <- m+sd(x)
   return(c(y=m,ymin=ymin,ymax=ymax))
}

Knee_height_EL_sexstrat<-ggplot(tmp, aes(x=sex,y=height_knee,fill=early_PC_split)) + 
  geom_violin()+ 
  stat_summary(fun.data=data_summary,position=position_dodge(0.9))+
  theme_bw(12) + 
  theme(legend.position="bottom", legend.title = element_text(size = 12),
    legend.text = element_text(size = 8), axis.text=element_text(color="black")) + 
  scale_color_manual(values=pal[1:2], labels=c("rural" = "rural \n(PC1 < 0)", "urban"= "urban \n(PC1 > 0)")) +
  guides(fill = guide_legend(title = "Early life urbanicity")) +
  scale_fill_manual(values=c("gray70", "white")) +
  xlab("") +
  ylab("Knee height");Knee_height_EL_sexstrat
```

### Make the forest plots 
```{r} 
##List of outcome variables - std
Body_Height <- c("height_standing_std", "height_hip_std", "height_knee_std")

Body_Press <- c("waist_circum_std", "weight_std", "BMI_std", "BRI_std", "body_fat_std", 
                "waist_hip_std","sys_bp2_std", "dias_bp2_std")

Blood_Bio <- c("total_chol_std", "ldl_std", "hdl_std", "triglyc_std")
all_outcomes <- c(Body_Height, Body_Press, Blood_Bio)

######
# early life
######

tmp<-as.data.frame(cbind(results_est[1:15,4],results_CI1[1:15,4],results_CI2[1:15,4],results_pval[1:15,4],p.adjust(results_pval[1:15,4],method='BH')))
tmp$outcome2<-all_outcomes
tmp$outcome<-factor(tmp$outcome2, levels=rev(all_outcomes), labels=rev(c("Standing height","Hip height","Knee height","Waist circumference", "Body weight", "BMI","BRI","Body fat percentage",  "Waist to hip ratio", "Systolic BP", "Diastolic BP", "Total cholesterol", "LDL", "HDL",  "Triglycerides")))

tmp$significant<-'ns'
tmp$significant[which(tmp$V4<0.05)]<-'p<0.05'
tmp$significant[which(tmp$V5<0.1)]<-'FDR<0.1'
tmp$significant <- factor(tmp$significant, levels = c("ns", "p<0.05", "FDR<0.1"))

dotCOLS = c("#89B9E1",'lightgrey',"#F8C17A")
barCOLS = c(pal[1],'darkgrey',pal[2])

EL_PC1_forest<-ggplot(tmp, aes(x=outcome, y=V1, ymin=V2, ymax=V3,col=significant,fill=significant)) + 
  geom_linerange(size=1,position=position_dodge(width = 0.5), show.legend=TRUE) +
  geom_hline(yintercept=0, lty=2) +
  geom_point(size=3, shape=21, colour="white", stroke = 0.5,position=position_dodge(width = 0.5), show.legend=TRUE) +
  ylim(-0.2, 0.41)+
  xlab("")+
  scale_fill_manual(values=c("ns"=barCOLS[2], "p<0.05"= barCOLS[3], "FDR<0.1"=barCOLS[1]), drop=F)+
  scale_color_manual(values=c("ns"=dotCOLS[2], "p<0.05"= dotCOLS[3], "FDR<0.1"=dotCOLS[1]), drop=F)+
  coord_flip() +
  theme_bw(12) + 
  theme(plot.title=element_text(hjust=0.5), legend.title=element_blank(), axis.text=element_text(color="black")) +
  labs(y='Effect size', title='Early life PC1');EL_PC1_forest

######
# current urbanicity
######

tmp<-as.data.frame(cbind(results_est[1:15,3],results_CI1[1:15,3],results_CI2[1:15,3],results_pval[1:15,3],p.adjust(results_pval[1:15,3],method='BH')))
tmp$outcome2<-all_outcomes
tmp$outcome<-factor(tmp$outcome2, levels=rev(all_outcomes), labels=rev(c("Standing height","Hip height","Knee height","Waist circumference", "Body weight", "BMI","BRI","Body fat percentage",  "Waist to hip ratio", "Systolic BP", "Diastolic BP", "Total cholesterol", "LDL", "HDL",  "Triglycerides")))
tmp$significant<-'ns'
tmp$significant[which(tmp$V4<0.05)]<-'p<0.05'
tmp$significant[which(tmp$V5<0.1)]<-'FDR<0.1'
tmp$significant <- factor(tmp$significant, levels = c("ns", "p<0.05", "FDR<0.1"))

dotCOLS = c("#89B9E1",'lightgrey',"#F8C17A")
barCOLS = c(pal[1],'darkgrey',pal[2])

Current_PC1_forest<-ggplot(tmp, aes(x=outcome, y=V1, ymin=V2, ymax=V3,col=significant,fill=significant)) + 
  geom_linerange(size=1,position=position_dodge(width = 0.5), show.legend=TRUE) +
  geom_hline(yintercept=0, lty=2) +
  geom_point(size=3, shape=21, colour="white", stroke = 0.5,position=position_dodge(width = 0.5), show.legend=TRUE) +
  ylim(-0.2, 0.41)+
  xlab("")+
  scale_fill_manual(values=c("ns"=barCOLS[2], "p<0.05"= barCOLS[3], "FDR<0.1"=barCOLS[1]), drop=F)+
  scale_color_manual(values=c("ns"=dotCOLS[2], "p<0.05"= dotCOLS[3], "FDR<0.1"=dotCOLS[1]), drop=F)+
  coord_flip() +
  theme_bw(12) + 
  theme(axis.text.y=element_blank(), plot.title=element_text(hjust=0.5), legend.title=element_blank())+
  labs(y='Effect size', title='Current urbanicity');Current_PC1_forest

######
# interaction effect
######
tmp<-as.data.frame(cbind(results_est[1:15,5],results_CI1[1:15,5],results_CI2[1:15,5],results_pval[1:15,5],p.adjust(results_pval[1:15,5],method='BH')))
tmp$outcome2<-all_outcomes
tmp$outcome<-factor(tmp$outcome2, levels=rev(all_outcomes), labels=rev(c("Standing height","Hip height","Knee height","Waist circumference", "Body weight", "BMI","BRI","Body fat percentage",  "Waist to hip ratio", "Systolic BP", "Diastolic BP", "Total cholesterol", "LDL", "HDL",  "Triglycerides")))
tmp$significant<-'ns'
tmp$significant[which(tmp$V4<0.05)]<-'p<0.05'
tmp$significant[which(tmp$V5<0.1)]<-'FDR<0.1'
tmp$significant <- factor(tmp$significant, levels = c("ns", "p<0.05", "FDR<0.1"))

dotCOLS = c("#89B9E1",'lightgrey',"#F8C17A")
barCOLS = c(pal[1],'darkgrey',pal[2])

Int_PC1_forest<-ggplot(tmp, aes(x=outcome, y=V1, ymin=V2, ymax=V3,col=significant,fill=significant)) + 
  geom_linerange(size=1,position=position_dodge(width = 0.5), show.legend=TRUE) +
  geom_hline(yintercept=0, lty=2) +
  geom_point(size=3, shape=21, colour="white", stroke = 0.5,position=position_dodge(width = 0.5), show.legend=TRUE) +
  ylim(-0.2, 0.41)+
  xlab("")+
  scale_fill_manual(values=c("ns"=barCOLS[2], "p<0.05"= barCOLS[3], "FDR<0.1"=barCOLS[1]), drop=F)+
  scale_color_manual(values=c("ns"=dotCOLS[2], "p<0.05"= dotCOLS[3], "FDR<0.1"=dotCOLS[1]), drop=F)+
  coord_flip() +
  theme_bw(12) + 
  theme(axis.text.y=element_blank(), plot.title=element_text(hjust=0.5), legend.title=element_blank(), axis.text=element_text(color="black"))+
  labs(y='Effect size', title='Interaction');Int_PC1_forest
```

         ############################################
         ######### Run sex stratefied model #########
         ############################################
         
```{r} 
all_outcomes <- c(Body_Height, Body_Press, Blood_Bio)
results_pval<-matrix(nrow=length(all_outcomes),ncol=10)
results_est<-matrix(nrow=length(all_outcomes),ncol=12)
results_CI1<-matrix(nrow=length(all_outcomes),ncol=8)
results_CI2<-matrix(nrow=length(all_outcomes),ncol=8)

OA_merged_F<-subset(OA_merged,sex=='Female')
OA_merged_M<-subset(OA_merged,sex=='Male')

for (i in 1:length(all_outcomes)){
OA_merged_F$tmp_outcome<-OA_merged_F[,which(names(OA_merged_F) %in% all_outcomes[i])]
x<-(lm( tmp_outcome ~ age_std + urb_score_std * Early_PC1_std ,data=OA_merged_F))
results_pval[i,1:4]<-summary(x)$coefficients[2:5,4]
results_est[i,1:4]<-summary(x)$coefficients[2:5,1]
results_CI1[i,1:4]<-confint(x)[2:5,1]
results_CI2[i,1:4]<-confint(x)[2:5,2]
results_est[i,5]<-nobs(x)
results_est[i,6]<-"female"
results_pval[i,5]<-"female"
}

for (i in 1:length(all_outcomes)){
OA_merged_M$tmp_outcome<-OA_merged_M[,which(names(OA_merged_M) %in% all_outcomes[i])]
x<-(lm( tmp_outcome ~ age_std + urb_score_std * Early_PC1_std ,data=OA_merged_M))
results_pval[i,6:9]<-summary(x)$coefficients[2:5,4]
results_est[i,7:10]<-summary(x)$coefficients[2:5,1]
results_CI1[i,5:8]<-confint(x)[2:5,1]
results_CI2[i,5:8]<-confint(x)[2:5,2]
results_est[i,11]<-nobs(x)
results_est[i,12]<-"male"
results_pval[i,10]<-"male"
}

colnames(results_est)[7:12] <- colnames(results_est)[1:6]
results_est<-as.data.frame(rbind(results_est[,1:6], results_est[,7:12]))
colnames(results_est)<-c('age','urb_score','early_PC1','int', 'nobs', 'sex')

colnames(results_pval)[6:10] <- colnames(results_pval)[1:5]
results_pval<-as.data.frame(rbind(results_pval[,1:5], results_pval[,6:10]))
colnames(results_pval)<-c('age','urb_score','early_PC1','int', 'sex')
results_est$outcome<-c(all_outcomes, all_outcomes)
results_pval$outcome<-c(all_outcomes, all_outcomes)
```

#export model results
```{r}
estimates<-pivot_longer(data=results_est, cols=1:4, names_to= "predictor", values_to="estimate")
pvalues<- pivot_longer(data=results_pval, cols=1:4, names_to= "predictor", values_to="pvalue")

merged<- merge(estimates, pvalues)
merged$FDR<- p.adjust(merged$pvalue,method='BH')

merged$outcome <- recode(merged$outcome,
  height_standing_std = "Standing height",
  height_hip_std = "Hip height",
  height_knee_std = "Knee height",
  waist_circum_std = "Waist circumference",
  weight_std = "Body weight",
  BMI_std = "BMI",
  BRI_std = "BRI",
  body_fat_std = "Body fat percentage",
  waist_hip_std = "Waist to hip ratio",
  sys_bp2_std = "Systolic BP",
  dias_bp2_std = "Diastolic BP",
  total_chol_std = "Total cholesterol",
  ldl_std = "LDL",
  hdl_std = "HDL",
  triglyc_std = "Triglycerides"
)

merged$predictor <- recode(merged$predictor,
  early_PC1 = "early life PC1",
  int = "interaction",
  urb_score = "current urbanicity")

merged$significant<-'ns'
merged$significant[which(merged$pvalue<0.05)]<-'p<0.05'
merged$significant[which(merged$FDR<0.1)]<-'FDR<0.1'
```


#Make forest plot of early life effects
```{r} 
all_outcomes <- c(Body_Height, Body_Press, Blood_Bio)
results_pval<-matrix(nrow=length(all_outcomes),ncol=8)
results_est<-matrix(nrow=length(all_outcomes),ncol=8)
results_CI1<-matrix(nrow=length(all_outcomes),ncol=8)
results_CI2<-matrix(nrow=length(all_outcomes),ncol=8)

OA_merged_F<-subset(OA_merged,sex=='Female')
OA_merged_M<-subset(OA_merged,sex=='Male')

for (i in 1:length(all_outcomes)){
OA_merged_F$tmp_outcome<-OA_merged_F[,which(names(OA_merged_F) %in% all_outcomes[i])]
x<-(lm(tmp_outcome ~ age_std + urb_score_std * Early_PC1_std ,data=OA_merged_F))
results_pval[i,1:4]<-summary(x)$coefficients[2:5,4]
results_est[i,1:4]<-summary(x)$coefficients[2:5,1]
results_CI1[i,1:4]<-confint(x)[2:5,1]
results_CI2[i,1:4]<-confint(x)[2:5,2]
}

for (i in 1:length(all_outcomes)){
OA_merged_M$tmp_outcome<-OA_merged_M[,which(names(OA_merged_M) %in% all_outcomes[i])]
x<-(lm( tmp_outcome ~ age_std + urb_score_std * Early_PC1_std ,data=OA_merged_M))
results_pval[i,5:8]<-summary(x)$coefficients[2:5,4]
results_est[i,5:8]<-summary(x)$coefficients[2:5,1]
results_CI1[i,5:8]<-confint(x)[2:5,1]
results_CI2[i,5:8]<-confint(x)[2:5,2]
}



tmpF<-as.data.frame(cbind(results_est[,3],results_CI1[,3],results_CI2[,3],results_pval[,3],p.adjust(results_pval[,3],method='BH')))
tmpF$outcome<-all_outcomes
tmpF$significant<-'ns'
tmpF$significant[which(tmpF$V4<0.05)]<-'p<0.05'
tmpF$significant[which(tmpF$V5<0.1)]<-'FDR<0.1'

tmpM<-as.data.frame(cbind(results_est[,7],results_CI1[,7],results_CI2[,7],results_pval[,7],p.adjust(results_pval[,7],method='BH')))
tmpM$outcome<-all_outcomes
tmpM$significant<-'ns'
tmpM$significant[which(tmpM$V4<0.05)]<-'p<0.05'
tmpM$significant[which(tmpM$V5<0.1)]<-'FDR<0.1'

tmpM$sex<-'male'
tmpF$sex<-'female'

tmpSPLIT<-rbind(tmpM,tmpF)

tmpSPLIT$outcome_format<-factor(tmpSPLIT$outcome, levels=rev(all_outcomes), labels=rev(c("Standing height","Hip height","Knee height","Waist circumference", "Body weight", "BMI","BRI","Body fat percentage",  "Waist to hip ratio", "Systolic BP", "Diastolic BP", "Total cholesterol", "LDL", "HDL",  "Triglycerides")))

tmpSPLIT$significant<-factor(tmpSPLIT$significant, levels=c("FDR<0.1", "p<0.05", "ns"))

dotCOLS = c("#89B9E1","#F8C17A", 'lightgrey')
barCOLS = c(pal[1],pal[2],'darkgrey')

EL_sexfacet<- ggplot(tmpSPLIT, aes(x=outcome_format, y=V1, ymin=V2, ymax=V3,col=significant,fill=significant)) + 
#specify position here
  geom_linerange(size=1,position=position_dodge(width = 0.5)) +
  geom_hline(yintercept=0, lty=2) +
#specify position here too
  geom_point(size=3, shape=21, colour="white", stroke = 0.5,position=position_dodge(width = 0.5)) +
  scale_fill_manual(values=barCOLS)+
  scale_color_manual(values=dotCOLS)+
  coord_flip() +
  theme_bw(12) + 
  theme(plot.title=element_text(hjust=0.5), strip.text=element_text(size=12), axis.text=element_text(color="black"), legend.position="bottom") + 
  labs(y='Effect size', x= "", title='Early life PC1') +facet_wrap(~sex); EL_sexfacet
```

          ##########################################################################
          ############### Run the regression test for DC versus PAR ################
          ##########################################################################
          
```{r} 
# List of outcome variables - std
OA_merged$ldl_std_rev<-OA_merged$ldl_std * -1

Body_Height <- c("height_standing_std", "height_hip_std", "height_knee_std")

Body_Press <- c("waist_circum_std", "weight_std", "BMI_std", "BRI_std", "body_fat_std", 
                "waist_hip_std","sys_bp2_std", "dias_bp2_std")

Blood_Bio <- c("total_chol_std", "ldl_std_rev", "hdl_std", "triglyc_std")

all_outcomes<-c(Body_Press,Blood_Bio)

regression_test1<-matrix(nrow=length(all_outcomes),ncol=7)
# e0 = early life (early)
# ed = difference between current and early (adult - early)

# Generate quadratic terms
# higher = rural/better 'quality'
OA_merged$e0 <- -1*(OA_merged$Early_PC1_std)
OA_merged$ed <- abs(-1*OA_merged$urb_score_std - OA_merged$e0)
OA_merged$e02 <- OA_merged$e0^2
OA_merged$ed2 <- OA_merged$ed^2
OA_merged$e0d <- OA_merged$e0 * OA_merged$ed

for (i in 1:length(all_outcomes)){
OA_merged$tmp_outcome<-OA_merged[,which(names(OA_merged) %in% all_outcomes[i])]*-1
# Basic quadratic regression
model <- lm(tmp_outcome ~ e0 + ed + e02 + ed2 + e0d + age_std + sex, data = OA_merged)

# Test for DC and PAR at mean of data
# Gather means of each input variable
tmp<-subset(OA_merged,!is.na(ed) & !is.na(tmp_outcome) )
m0 <- mean(tmp$e0,na.rm=T)
md <- mean(tmp$ed,na.rm=T)

# Test for DC 
dc_est <- as.numeric(coef(model)[2]) + (2 * as.numeric(coef(model)[4]) * m0) + (as.numeric(coef(model)[6]) * md)
dc_n<-nobs(model)
pval_dc <- as.numeric(nlWaldtest(model,c(paste("b[2] + 2*b[4]*",m0," + b[6]*",md))) [4])

regression_test1[i,1]<- pval_dc
regression_test1[i,2]<- dc_est
regression_test1[i,7]<-dc_n

# Test for PAR 
# similar result using ed for the p-value
par_est <- as.numeric(coef(model)[3]) + (2 * as.numeric(coef(model)[5]) * md) + (as.numeric(coef(model)[6]) * m0)
pval_par <- as.numeric(nlWaldtest(model,c(paste("b[3] + 2*b[5]*",md," + b[6]*",m0))) [4])

regression_test1[i,4]<- pval_par
regression_test1[i,5]<- par_est

}
regression_test1[,3]<-p.adjust(regression_test1[,1],method='BH')
regression_test1[,6]<-p.adjust(regression_test1[,4],method='BH')

regression_test2<-as.data.frame(rbind(regression_test1[,c(1:3,7)],regression_test1[,4:7]))
regression_test2$outcome<-c(all_outcomes,all_outcomes)
regression_test2$test<-c(rep('DC',length(all_outcomes)),rep('PAR',length(all_outcomes)))
regression_test2$sig<-'ns'
regression_test2$sig[which(regression_test2$V1<0.05 & regression_test2$V2>0 & regression_test2$test=='DC')]<-'p<0.05'
regression_test2$sig[which(regression_test2$V1<0.05 & regression_test2$V2<0 & regression_test2$test=='PAR')]<-'p<0.05'
regression_test2$sig[which(regression_test2$V3<0.1 & regression_test2$V2>0 & regression_test2$test=='DC')]<-'FDR<0.1'
regression_test2$sig[which(regression_test2$V3<0.1 & regression_test2$V2<0 & regression_test2$test=='PAR')]<-'FDR<0.1'
regression_test2$outcome_format<-factor(regression_test2$outcome, levels=rev(c("weight_std", "body_fat_std", "BMI_std", "waist_circum_std", "waist_hip_std", "BRI_std", "sys_bp2_std", "dias_bp2_std", "total_chol_std", "hdl_std", "ldl_std_rev",  "triglyc_std", "glucose_std")), labels=rev(c("Body weight", "Body fat percentage", "BMI", "Waist circumference", "Waist to hip ratio", "BRI", "Systolic BP", "Diastolic BP", "Total cholesterol", "HDL", "LDL",  "Triglycerides", "Glucose")))
regression_test2$sig<-factor(regression_test2$sig, levels=c("ns", "p<0.05", "FDR<0.1"))
colnames(regression_test2) <- c("pvalue", "estimate", "FDR", "sample_size", "outcome", "test", "sig", "outcome_format")

DCvPAR_dotplot<-ggplot(regression_test2, aes(x=outcome_format, y=estimate,color=sig)) +
  geom_segment(aes(x=outcome_format, xend=outcome_format, y=0, yend=estimate))  +
  geom_point() +
  theme_bw(13) +
  theme(strip.text.x=element_text(size=14), axis.text.x=element_text(angle=45, hjust=0.5, vjust=0.5), legend.title=element_blank(), legend.position="top") +
  coord_flip() +
  xlab("") + ylab("Estimate") +facet_wrap(~test) +
  scale_color_manual(values=c("FDR<0.1"=pal[1], "p<0.05"=pal[2], "ns"="darkgray")); DCvPAR_dotplot
```

### Plot two examples
```{r} 
# example
A<-ggplot(OA_merged, aes(x=e0, y= body_fat)) +
  geom_text(x=0.5, y=55, size=4.5, label=expression(paste("p = 3.2 x ", 10^-14))) +
  geom_point(size=1) + stat_smooth(method='lm', color=pal[1])+
  theme_bw(13) + 
  theme(plot.title=element_text(size=14, hjust=0.5)) +
  xlab('') +
  ylab("Body fat percentage") +
  ggtitle("");A


B<-ggplot(OA_merged, aes(x= (-1 * urb_score_std) - e0, y=body_fat)) +
  geom_point(size=1) + 
  stat_smooth(method='lm', formula = y ~ x + I(x^2), color="darkgray")+
  theme_bw(13) + 
  theme(plot.title=element_text(size=14, hjust=0.5))+
  xlab('') + 
  geom_text(x=3, y=55, size=4.5, label="p = 0.98") +
  ylab("Body fat percentage") +
  ggtitle(""); B
```

#sessionInfo()
```{r, eval=FALSE}
R version 4.5.1 (2025-06-13)
Platform: aarch64-apple-darwin20
Running under: macOS Tahoe 26.0.1

Matrix products: default
BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib 
LAPACK: /Library/Frameworks/R.framework/Versions/4.5-arm64/Resources/lib/libRlapack.dylib;  LAPACK version 3.12.1

locale:
 en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

time zone: America/Chicago
tzcode source: internal

attached base packages:
 stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
nationalparkcolors_0.1.0 nlWaldTest_1.1.3         ggmap_4.0.2              ggpubr_0.6.2            
  mice_3.19.0              lubridate_1.9.4          forcats_1.0.1            stringr_1.6.0           
  dplyr_1.1.4              purrr_1.2.1              readr_2.1.6              tidyr_1.3.2             
 tibble_3.3.1             ggplot2_4.0.1            tidyverse_2.0.0         

loaded via a namespace (and not attached):
  tidyselect_1.2.1   farver_2.1.2       S7_0.2.1           bitops_1.0-9       fastmap_1.2.0     
  digest_0.6.39      rpart_4.1.24       timechange_0.3.0   lifecycle_1.0.5    survival_3.8-6    
 magrittr_2.0.4     compiler_4.5.1     rlang_1.1.7        tools_4.5.1        yaml_2.3.12       
 knitr_1.51         ggsignif_0.6.4     labeling_0.4.3     curl_7.0.0         plyr_1.8.9        
 RColorBrewer_1.1-3 abind_1.4-8        withr_3.0.2        nnet_7.3-20        grid_4.5.1        
 jomo_2.7-6         scales_1.4.0       iterators_1.0.14   MASS_7.3-65        cli_3.6.5         
 rmarkdown_2.30     ragg_1.5.0         reformulas_0.4.3.1 generics_0.1.4     otel_0.2.0        
 rstudioapi_0.18.0  httr_1.4.7         tzdb_0.5.0         minqa_1.2.8        splines_4.5.1     
 vctrs_0.7.0        boot_1.3-32        glmnet_4.1-10      Matrix_1.7-4       carData_3.0-5     
 car_3.1-3          hms_1.1.4          mitml_0.4-5        rstatix_0.7.3      Formula_1.2-5     
 systemfonts_1.3.1  jpeg_0.1-11        foreach_1.5.2      glue_1.8.0         nloptr_2.2.1      
 pan_1.9            codetools_0.2-20   cowplot_1.2.0      stringi_1.8.7      shape_1.4.6.1     
 gtable_0.3.6       lme4_1.1-38        pillar_1.11.1      htmltools_0.5.9    R6_2.6.1          
 textshaping_1.0.4  Rdpack_2.6.4       evaluate_1.0.5     lattice_0.22-7     rbibutils_2.4     
 png_0.1-8          backports_1.5.0    broom_1.0.11       Rcpp_1.1.1         gridExtra_2.3     
 nlme_3.1-168       mgcv_1.9-4         xfun_0.56          pkgconfig_2.0.3  
```

